Pythran: Enabling Static Optimization of Scientific Python Programs

### 2014-01-28 doc/papers/iop2014/paper.tex
Review IOP 2014 paper

    In addition to these compilation steps, Pythran provides a C++ runtime library that
    leverages the C++ STL to provide generic containers, and the Numeric
    Template Toolbox (NT2) for Numpy support. It takes advantage of modern C++11
    features such as variadic templates, type inference, move semantics and
    perfect forwarding, as well as classical ones such as expression templates.

    The input code remains compatible with the Python interpreter, and output
    code is generally as efficient as the annotated Cython equivalent, if not
    more, without the backward compatibility loss of Cython. Numpy expressions
    run faster than when compiled with numexpr, or numba, without any change to
    the original code.

@@

Several tools have been proposed by an active community to fill the performance
gap met when running these computation-intensive piece of code, either through
static compilation or Just In Time (JIT) compilation.

@@

To the notable exception of PyPy, these compilers do not apply any of the
static optimization techniques that have been known for decades and
successfully applied to statically compiled language such as C or C++.
Translators to statically compiled languages do take advantage of them
indirectly, but the quality of generated code may prevent advanced
optimizations, such as vectorization, while they are available at higher level,
i.e. at the Python level. Taking into account the specificities of the Python
language can unlock many new transformations. For instance, PyPy automates the
conversion of the \texttt{range} builtin into \texttt{xrange} through the use
of a dedicated structure called \texttt{range-list}.

@@

Unlike existing alternatives, Pythran does not solely perform static typing of
Python programs. It also performs various compiler optimizations such as
detection of pure functions, temporary allocation removal or constant folding.
These transformations are backed up by code analysis such as aliasing,
inter-procedural memory effect computations or use-def chains.

The article is structured as follows: Section~\ref{sec:infrastructure}
introduces the Pythran compiler compilation flow and internal representation.
Section~\ref{sec:analysis} presents several code analysis while
Section~\ref{sec:optimizations} focuses on code optimizations.
Section~\ref{sec:backend} presents back-end optimizations for Numpy
expressions. Section~\ref{sec:openmp}  briefly introduces OpenMP-like
annotations for explicit parallelization of Python programs and
section~\ref{sec:benchmarks} presents the performance obtained on a few
synthetic benchmarks and concludes.

@@

\begin{lstlisting}
def dot(l0, l1):
    return sum(x*y for x,y in zip(l0,l1))
\end{lstlisting}

@@

\begin{lstlisting}[language=c++]
template<class T0, class T1>
    auto dot(T0&& l0, T1&& l1) -> decltype(/* skipped */)
    {
        return pythonic::sum(
            pythonic::map(
                operator_::multiply(),
                    pythonic::zip(
                        std::forward<T0>(l0),
                        std::forward<T1>(l1))
            )
        );
    }
\end{lstlisting}

Although far more verbose than the Python version, the C++ version also uses a
form of structural typing: the only assumption these two versions make are that
\texttt{l0} and \texttt{l1} are iterable, their content can be multiplied and the result of
the multiplication is accumulatable.

@@

The second step only consists in the instantiation of the top-level functions of the
module, using user-provided signatures. Template instantiation then triggers the
different correctly typed instantiations for all functions written in the
module. Note that the user only needs to provide the type of the functions
exported outside the module. The possible types of all internal functions are
then inferred from the call sites.

The last step involves a template library, called \texttt{pythonic} that contains a
polymorphic implementation of many functions from the Python standard library
in the form of C++ template functions. Several optimizations, most notably
expression template, are delegated to this library. Pythran relies on the
C++11\cite{isocxx11} language, as it makes heavy use of recent features such as
move semantics, type inference through \texttt{decltype(\dots)} and variadic templates.
As a consequence it requires a compatible C++ compiler for the native code
generation. Boost.Python\cite{boostpython2007} is involved for the automatic
Python-to-C++ boilerplate.  Generated C++ code is compatible with g++ 4.7.2 and
clang++ 3.2.

It is important to note that all Pythran analyses and optimizations are
type-agnostic, i.e. they do not assume any type for the variables manipulated
by the program. Type specialization is only done in the back-end, right before
native code generation. Said otherwise, the Pythran compiler analyzes
polymorphic functions and turns Python modules into C++ meta-programs.

Figure\ref{fig:pythran-compiler} summarizes the compilation flow and the involved
tools.

@@

A code analysis is a function that takes a part of the IR (or the whole
module's IR) as input and returns aggregated high-level information. For
instance, a simple Pythran analysis called \texttt{Identifiers} gathers the set
of all identifiers used throughout the program. This information is later used
when the creation of new identifiers is required so that no conflict occurs
with existing ones.

@@

The fact that all paths finish by a \textbf{def} indicates that the last
assignment can be removed (but not necessarily its right hand part that could
have a side-effect). The fact that the last \textbf{use} is always preceded by
a \textbf{def}taht only point to it indicates that these \textbf{def}s can
safely be renamed from \texttt{a} to \texttt{a}' (thus performing scalar
renaming).

@@

One of the benefits of translating Python code to C++ code is that it removes
most of the dynamic lookups. It also unveils all the optimizations available at
C++ level. For instance, a function call is quite costly in Python, which
advocates in favor of using inlining. This transformation comes at no cost when
using C++ as the back-end language, as the C++ compiler does it.

@@

\begin{lstlisting}
def valid_conversion(n):
    # this map can be converted to imap
    l = map(math.cos, range(n))
    return sum(l) # sum iterates once on its input

def invalid_conversion(n):
    # this map cannot be converted to imap
    l = map(math.cos, range(n))
    l[0] = 1  # invalid assignment
    return sum(l) + max(l) # sum iterates once
\end{lstlisting}

@@

Pythran makes this directives available at the Python level through string
instructions. The semantic is roughly similar to the original semantics,
assuming that all variables have function level scope.

@@

\begin{lstlisting}
def pi_estimate(darts):
    hits = 0
    #omp parallel for reduction(+:hits)"
    for i in xrange(darts):
        x,y = random(), random()
        dist = sqrt(pow(x, 2) + pow(y, 2))
        if dist <= 1.0:
            hits += 1.0
    pi = 4 * (hits / DARTS)
    return pi
\end{lstlisting}

@@

All benchmarks presented in this section are ran on an hyper-threaded quad-core
i7, using examples shipped along Pythran sources, available at
https://github.com/serge-sans-paille/pythran in the \texttt{pythran/test/cases}
directory. The Pythran version used is the \texttt{HEAD} of the \texttt{scipy2013} branch,
ShedSkin 0.9.2, PyPy 2.0 compiled with the \texttt{-jit} flag, CPython 2.7.3, Cython
0.19.1 and Numexpr 2.0.1. All timings are made using the \texttt{timeit} module,
taking the best of all runs. All C++ codes are compiled with g++ 4.7.3, using
the tool default compiler option, generally \texttt{-O2} plus a few optimizing flags
depending on the target.

@@

Pystone is a Python translation of whetstone, a famous floating point number
benchmarks that dates back to Algol60 and the 70's. Although non representative
of real applications, it illustrates the general performance of floating point
number manipulations. Table \ref{tbl:pystone}  illustrates the benchmark
result for CPython, PyPy, ShedSkin and Pythran, using an input value of
\texttt{10**3}. Note that the original version has been updated to replace the user
class by a function call.

@@

Nqueen is a benchmark extracted from the former Unladen Swallow\footnote{\url{http://code.google.com/p/unladen-swallow/}} project. It
is particularly interesting as it makes an intensive use of non-trivial
generator expressions and integer sets. Table\ref{tbl:nqueen} illustrates
the benchmark results for CPython, PyPy, ShedSkin and Pythran. The code had to
be slightly updated to run with ShedSkin because type inference in ShedSkin does
not support mixed scalar and \textbf{None} variables. The input value is \texttt{9}.

@@

Table\ref{tbl:hyantes} illustrates the benchmark result for CPython, PyPy,
ShedSkin and Pythran, when using lists as the data container. The output window
used is \texttt{100x100}.

@@

The speed ups are not amazing for a numerical application. there are two
reasons for this poor speedups. First, the \texttt{hyantes} benchmark makes heavy
usage of trigonometric functions, and there is not much gain there. Second, and
most important, the benchmark produces a big 2D array stored as a list of list,
so the application suffers from the heavy overhead of converting them from C++
to Python. Running the same benchmark using Numpy arrays as core containers
confirms this assumption, as illustrated by Table \ref{tbl:np-hyantes}. This
table also demonstrates the benefits of manual parallelization using OpenMP.

@@

Table \ref{tbl:arc-distance} illustrates the benchmark result for CPython,
Cython, Numexpr and Pythran, using random input arrays of \texttt{10**6} elements.
Table \ref{tbl:arc-distance-2} details the Pythran performance. Cython code
is written using the \texttt{parallel.prange} feature and compiled with
\texttt{-fopenmp -O2 -march=native}.

@@

\begin{table}
    \centering
    \begin{tabular}{|l|c|c|c|c|}
    \hline
    Tool & Pythran (raw) & Pythran (+AVX) & Pythran (+OMP) & Pythran (full)  \\
    \hline
    Timing & 186.3ms     &    75.4ms      &    41.1ms      &  17.1ms         \\
    \hline
    Speedup&    x1.03      &    x2.54       &    x4.67       &  x11.23         \\
    \hline
\end{tabular}
\caption{Benchmarking result on the arc distance kernel, Pythran details.}
\label{tbl:arc-distance-2}
\end{table}

It shows a small benefit from using expression templates on their own, most
certainly because the loop control overhead is negligible in front of the
trigonometric functions. It gets a decent x2.5 speed-up when using AVX over not
using it. The benefit of OpenMP, although related to the number of cores, makes
a whole speedup greater than x11 over the original Numpy version, without
changing the input code. Quite the opposite, Numexpr requires rewriting the input
and does not achieve the same level of performance as Pythran when OpenMP and
AVX are combined.

@@

Although Pythran focuses on a subset of Python and its standard library, many
optimizations opportunities are still possible. Using as Domain Specific
Language(DSL) approach, one could use rewriting rules to optimize several
Python idioms. For instance, \texttt{len(set(x))} could lead to an optimized
\texttt{count\_uniq} that would iterate only once on the input sequence.

>>>>

    In addition to these compilation steps, Pythran provides a C++ runtime library that
    leverages the C++ STL to provide generic containers, and the Numeric
    Template Toolbox (NT2) for Numpy support. It takes advantage of modern C++11
    features such as variadic templates, type inference, move semantics and
    perfect forwarding, as well as classical ones such as expression templates.
    % FIXME Expression template is not a "feature" of the language IMO

    The input code remains compatible with the Python interpreter, and output
    code is generally as efficient as the annotated Cython equivalent, if not
    more, without the backward compatibility loss of Cython. Numpy expressions
    run faster than when compiled with numexpr or numba, without any change to
    the original code.

@@

Several tools have been proposed by an active community to fill the performance
gap encountered when running these computation-intensive piece of code, either
through static compilation or Just In Time (JIT) compilation.

@@

To the notable exception of PyPy, these compilers do not apply any of the
static optimization techniques that have been known for decades and
% FIXME And Numba? It generates LLVM, which uses static optimization techniques
successfully applied to statically compiled language such as C or C++.
Translators to statically compiled languages do take advantage of them
indirectly, but the quality of generated code may prevent advanced
optimizations, such as vectorization, while they are available at higher level,
i.e. at the Python level. Taking into account the specificities of the Python
language can unlock many new transformations. For instance, PyPy automates the
conversion of the \texttt{range} builtin into \texttt{xrange} through the use
of a dedicated structure called \texttt{range-list}.

@@

Unlike existing alternatives, Pythran does not solely perform static typing of
Python programs. It also performs various compiler optimizations such as
detection of pure functions, temporary allocation removal or constant folding.
These transformations are backed up by code analyses such as aliasing,
inter-procedural memory effect computations or use-def chains.

The article is structured as follows: Section~\ref{sec:infrastructure}
introduces the Pythran compiler compilation flow and internal representation.
Section~\ref{sec:analysis} presents several code analyses while
Section~\ref{sec:optimizations} focuses on code optimizations.
Section~\ref{sec:backend} presents back-end optimizations for Numpy
expressions. Section~\ref{sec:openmp}  briefly introduces OpenMP-like
annotations for explicit parallelization of Python programs and
Section~\ref{sec:benchmarks} presents the performance obtained on a few
synthetic benchmarks and concludes.

@@

\begin{lstlisting}
def dot(L0, L1):
    return sum(x*y for x,y in zip(L0,L1))
\end{lstlisting}

@@

\begin{lstlisting}[language=c++]
template<class T0, class T1>
    auto dot(T0&& L0, T1&& L1) -> decltype(/* skipped */)
    {
        return pythonic::sum(
            pythonic::map(
                operator_::multiply(),
                    pythonic::zip(
                        std::forward<T0>(L0),
                        std::forward<T1>(L1))
            )
        );
    }
\end{lstlisting}

Although far more verbose than the Python version, the C++ version also uses a
form of structural typing: the only assumption these two versions make are that
\texttt{L0} and \texttt{L1} are iterable, their content can be multiplied and the result of
the multiplication is accumulatable.

@@

The second step only consists in the instantiation of the top-level functions of
the module, using user-provided signatures. Template instantiation then triggers
the different correctly typed instantiations for all functions written in the
module. Note that the user only needs to provide the type of the functions
exported outside the module. The possible types of all internal functions are
then inferred from the call sites. Note that this complex step leverage the 
C++11 compiler capabilities: Pythran generates only one line using user-provided
signature to trigger the template instanciation by the C++ compiler.

The last step involves a template library, called \texttt{pythonic} that 
contains a polymorphic implementation of many functions from the Python standard
library in the form of C++ template functions. Several optimizations, most
notably using expression template, are delegated to this library. Pythran relies
on the C++11\cite{isocxx11} language, as it makes heavy use of recent features 
such as move semantics, type inference through \texttt{decltype(\dots)} and 
variadic templates.
As a consequence it requires a C++11 compatible compiler for the native code
generation. Boost.Python\cite{boostpython2007} is involved for the automatic
Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
4.7.2 and clang++ 3.2.

It is important to note that all Pythran analyses and optimizations are
type-agnostic, i.e. they do not assume any type for the variables manipulated
by the program. The generated C++ template is still type agnostic, the specialization occurs only by generating the template instanciation.
Said otherwise, the Pythran compiler analyzes polymorphic functions and turns
Python modules into C++ meta-programs.

Figure~\ref{fig:pythran-compiler} summarizes the compilation flow and the 
involved tools.

@@

A code analysis is a function that takes a part of the IR (or the whole
module's IR) as an input and returns aggregated high-level information. For
instance, a simple Pythran analysis called \texttt{Identifiers} gathers the set
of all identifiers used throughout the program. This information is later used
when the creation of new identifiers is required so that no conflict occurs
with existing ones.

@@

The fact that all paths finish by a \textbf{def} indicates that the last
assignment can be removed (but not necessarily its right hand part that could
have a side-effect). The fact that the last \textbf{use} is always preceded by
a \textbf{def} that only point to it indicates that these \textbf{def}s can
safely be renamed from \texttt{a} to \texttt{a}' (thus performing scalar
renaming).
% FIXME The scalar renaming part is unclear

@@

One of the benefits of translating Python code to C++ code is that it removes
most of the dynamic lookups. It also unveils all the optimizations available at
C++ level. For instance, a function call is quite costly in Python, which
advocates in favor of using inlining. This transformation comes at no cost when
using C++ as the back-end language, as the C++ compiler performs this 
transformation triggered by fine tuned heuristics.

@@

\begin{lstlisting}
def valid_conversion(n):
    # this map can be converted to imap
    L = map(math.cos, range(n))
    return sum(l) # sum iterates once on its input

def invalid_conversion(n):
    # this map cannot be converted to imap
    L = map(math.cos, range(n))
    L[0] = 1  # invalid assignment if converted to imap
    return sum(L) + max(L) # two iterations over L prevents imap transformation
\end{lstlisting}

@@

Pythran makes this directives available at the Python level through string
instructions. The semantic is roughly similar to the semantics defined by the
standard, assuming that all variables have function-level scope.

@@

\begin{lstlisting}
def pi_estimate(darts):
    hits = 0
    "#omp parallel for reduction(+:hits)"
    for i in xrange(darts):
        x,y = random(), random()
        dist = sqrt(pow(x, 2) + pow(y, 2))
        if dist <= 1.0:
            hits += 1.0
    pi = 4 * (hits / DARTS)
    return pi
\end{lstlisting}

@@

All benchmarks presented in this section are run on an hyper-threaded quad-core
i7, using examples shipped along with Pythran sources, available at
https://github.com/serge-sans-paille/pythran in the \texttt{pythran/test/cases}
directory. The Pythran version used is the \texttt{HEAD} of the \texttt{scipy2013} branch,
ShedSkin 0.9.2, PyPy 2.0 compiled with the \texttt{-jit} flag, CPython 2.7.3, Cython
0.19.1 and Numexpr 2.0.1. All timings are made using the \texttt{timeit} module,
taking the best of all runs. All C++ codes are compiled with g++ 4.7.3, using
the tool default compiler option, generally \texttt{-O2} plus a few optimizing flags
depending on the target.

@@

Pystone is a Python translation of Whetstone, a famous floating point number
benchmarks that dates back to Algol60 and the 70's. Although non representative
of real applications, it illustrates the general performance of floating point
number manipulations. Table~\ref{tbl:pystone}  illustrates the benchmark
result for CPython, PyPy, ShedSkin and Pythran, using an input value of
\texttt{10**3}. Note that the original version has been updated to replace the user
class by a function call.

@@

Nqueen is a benchmark extracted from the former Unladen Swallow\footnote{\url{http://code.google.com/p/unladen-swallow/}} project. It
is particularly interesting as it makes an intensive use of non-trivial
generator expressions and integer sets. Table~\ref{tbl:nqueen} illustrates
the benchmark results for CPython, PyPy, ShedSkin and Pythran. The code had to
be slightly updated to run with ShedSkin because type inference in ShedSkin does
not support mixed scalar and \textbf{None} variables. The input value is \texttt{9}.

@@

Table~\ref{tbl:hyantes} illustrates the benchmark result for CPython, PyPy,
ShedSkin and Pythran, when using lists as the data container. The output window
used is \texttt{100x100}.

@@

The speed ups are not amazing for a numerical application. there are two
reasons for this poor speedups. First, the \texttt{hyantes} benchmark makes heavy
usage of trigonometric functions, and there is not much gain there. Second, and
most important, the benchmark produces a big 2D array stored as a list of list,
so the application suffers from the heavy overhead of converting them from C++
to Python. Running the same benchmark using Numpy arrays as core containers
confirms this assumption, as illustrated by Table~\ref{tbl:np-hyantes}. This
table also demonstrates the benefits of manual parallelization using OpenMP.

@@

Table~\ref{tbl:arc-distance} illustrates the benchmark result for CPython,
Cython, Numexpr and Pythran, using random input arrays of \texttt{10**6} elements.
Table~\ref{tbl:arc-distance-2} details the Pythran performance. Cython code
is written using the \texttt{parallel.prange} feature and compiled with
\texttt{-fopenmp -O2 -march=native}.

@@

\begin{table}
    \centering
    \begin{tabular}{|l|c|c|c|c|}
    \hline
    Tool & Pythran (raw) & Pythran (+AVX) & Pythran (+OMP) & Pythran (full)  \\
    \hline
    Timing & 186.3ms     &    75.4ms      &    41.1ms      &  17.1ms         \\
    \hline
    Speedup&    x1.03      &    x2.54       &    x4.67       &  x11.23         \\
    \hline
\end{tabular}
\caption{Detailed benchmarking result on the arc distance kernel for Pythran.}
\label{tbl:arc-distance-2}
\end{table}

It shows a small benefit from using expression templates on their own, most
certainly because the loop control overhead is negligible in front of the
trigonometric functions. It gets a decent x2.5 speed-up when using AVX over not
using it. The benefit of OpenMP, although related to the number of cores, makes
a whole speedup greater than x11 over the original Numpy version, without
changing the input code. Almost on the opposite, Numexpr requires rewriting the 
input and does not achieve the same level of performance as Pythran when OpenMP
and AVX are combined.

@@

Although Pythran focuses on a subset of Python and its standard library, many
optimizations opportunities are still possible. Using a Domain Specific
Language(DSL) approach, one could use rewriting rules to optimize several
Python idioms. For instance, \texttt{len(set(x))} could lead to an optimized
\texttt{count\_uniq} that would iterate only once on the input sequence.
